---
title: "null"
csl: plos.csl
fontsize: 10pt
output:
  pdf_document:
    includes:
      in_header: latex/header.tex
    keep_tex: yes
  html_document:
    df_print: paged
  word_document: default
link-citations: yes
geometry: left=2cm, right=2cm, top=2cm, bottom=3cm, footskip = .5cm
always_allow_html: true
subparagraph: yes
bibliography: SOE2020.bib
---

```{r setup, include=FALSE}

# library(tint)
# # invalidate cache when the package version changes
# knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
# options(htmltools.dir.version = FALSE)

#Default Rmd options
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      dev = "cairo_pdf",
                      warning = FALSE,
                      fig.width = 4,
                      fig.asp = 0.45,
                      fig.align = 'center'
                      ) #allows for inserting R code into captions

#Plotting and data libraries
#remotes::install_github("noaa-edab/ecodata@0.1.0") #change to 2020 ecodata version for release
library(tidyverse)
library(tidyr)
library(ecodata)
library(here)
library(kableExtra)

```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-setup.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-GIS-setup.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_NE.Rmd-setup.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-setup.R")}
```

\setcounter{page}{3}

# Characterizing ecosystem change for fishery management

The Northeast shelf ecosystem is changing which is affecting the services that the ecosystem provides. To illustrate how multiple factors are driving change in this complex ecosystem we are using three overarching concepts: multiple system drivers, regime shifts, and ecosystem reorganization. Physical, chemical, biological, and human factors comprise the **multiple system drivers** that influence the quality and distribution of habitat in the ocean, which affects the amount and diversity of fish in the ocean. These environmental drivers combined with multiple social and economic drivers influence the range of fishing opportunities and the seafood, recreation, and other ecosystem services we derive from the ocean. 

```{r, out.width='72%'}

knitr::include_url("https://github.com/NOAA-EDAB/presentations/raw/master/docs/EDAB_images/multipledrivers.png")

```

Changes in those drivers can lead to **regime shifts** — large, abrupt and persistent changes in the structure and function of an ecosystem. Regime shifts and changes in how the multiple system drivers interact can result in **ecosystem reorganization** as species and humans respond and adapt to the new environment. 

```{r, out.width='72%'}

knitr::include_url("https://github.com/NOAA-EDAB/presentations/raw/master/docs/EDAB_images/RegimeReorg.png")
```

# Report Structure

The purpose of this report is to synthesize ecosystem information to better meet fishery management objectives. The major messages of the report are synthesized in the 2-page summary. The information in this report is organized into two sections; performance against ecosystem-level management objectives (Table \ref{tab:management-objectives}), and external considerations for fishery management (Climate change and Other Ocean Uses).  

Detailed technical methods documentation^[https://NOAA-EDAB.github.io/tech-doc] and indicator data^[https://github.com/NOAA-EDAB/ecodata] are available online. The details of standard figure formatting (Fig. \ref{fig:docformat}a), categorization of fish and invertebrate species into feeding groups (Table \ref{tab:species-groupings}), and definitions of ecological production units (EPUs, including Georges Bank, GB and the Gulf of Maine, GOM; Fig. \ref{fig:docformat}b) are provided at the end of the document. 

```{r management-objectives}

mng_obj <- data.frame("Objective Categories" = c("Seafood Production",
                                                 "Profits","Recreation",
                                                 "Stability","Social & Cultural",
                                                 "Protected Species",
                                                 "Biomass","Productivity",
                                                 "Trophic structure","Habitat"),
"Indicators reported here" = c("Landings by feeding guild","Revenue decomposed to price and volume",
                               "Days fished; recreational catch",
                               "Diversity indices (fishery and species)",
                               "Community engagement/reliance status",
                               "",
                               "Biomass or abundance by feeding guild from surveys",
                               "Condition and recruitment of managed species, Primary productivity",
                               "Relative biomass of feeding guilds, Zooplankton",
                               "Estuarine and offshore habitat conditions"))

knitr::kable(mng_obj, linesep = "",
      col.names = c("Objective Categories","Indicators reported here"),
      caption = "Established ecosystem-scale objectives in the Mid-Atlantic Bight",
      #align = 'c',
      booktabs = T) %>%
 # kable_styling(latex_options = "hold_position", "scale_down") %>%
 # column_spec(c(2), width = c("25em")) %>%
  row_spec(0, bold = TRUE) %>%
 # group_rows("Provisioning/Cultural", 1,4) %>%
 # group_rows("Supporting/Regulating", 5,9)
  pack_rows("Provisioning/Cultural Services", 1,6) %>%
  pack_rows("Supporting/Regulating Services", 7,10)


```
 
# Performance relative to fishery management objectives

In this section, we examine indicators related to broad, ecosystem-level fishery management objectives. We also provide hypotheses about *why* we are seeing these trends by identifying interacting drivers (other indicators in this report), and where possible, any regime shifts or potential changes in ecosystem structure. Identifying multiple drivers, regime shifts, and potential changes to ecosystem structure, as well as mapping most vulnerable resources, can help managers determine whether we can do anything differently to meet objectives and how to prioritize for upcoming issues/risks. 

## Seafood Production 

### Indicator: Landings

Total commercial landings within the GOM and NEFMC's managed landings in both regions continue to trend down. 

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-comdat-total-landings.R"), fig.width = 7.5}
```

The downward trend on Georges Bank is caused by a decline in the piscivores and benthos groups and slightly offset buy increases in the planktivores.  The trend in the GOM is seen is the piscivores, planktivores, and benthivores groups.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-comdat-commercial-landings.R"), fig.asp=1.5}
```

Recreational landings are also down overall in New England.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-recdat-landings.R")}

```

Recreational shark landings show increased catch of pelagic and prohibited sharks over the last couple of years.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-rec_hms.R")}

```

Aquaculture production is not yet included in total seafood landings, but we are working towards that in future reports. Available aquaculture production of oysters for a subset of New England states is showing an increase in the production per acre as well as the number of acres leased.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-aquaculture.R"), fig.asp = 1.5}

```

### Why

The decline in seafood production in the region could be cause by several factors.  Thresholds for ecosystem-level fishing based on system production characteristics have been proposed and are applied here for GB and GOM.  Based on either the ratio of total catch to total primary production (Fogarty), or total catch per unit area (Ryther).  Based on the proposed thresholds, there is a possibility that marginal ecosystem overfishing is occurring. This is further corroborated by the nine stocks below $B_{MSY}$ while status relative to $B_{MSY}$ could not be assessed for 14 additional stocks.  

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-fogarty.R"), fig.width = 7.5}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-ryther.R"), fig.width = 7.5}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-stock-status.R"), fig.width = 7.5, fig.asp = 0.5}
```

Fisheries account for a lower proportion of primary production removal now than in the past for GB while accounting for an average portion within the GOM.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-ppr.R"), fig.width = 7.5}
```

Even with potential ecosystem overfishing, aggregate biomass trends have been relatively stable in both regions, the lone execption being the benthivores group driven by recent haddock recruitment.  Even that group appears to be returning to average (although many surveys were not completed in 2020). Overall biomass does not appear to drive declining landings.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_NE.Rmd-aggregate-biomass.R"), fig.width=8, fig.asp = 0.75}
```

Commercial landings decline:  
  
Recreational landings decline: 
- Recreational shark landings declining due to management (previously less regulated)
- Most recent lows in all other recreational fish could still be the change in MRIP survey, calibration not quite lining up. Limits actually increased here so less likely management  
- Won't know till we get more data points  
- 80s/90s vs recent rec: could also be survey instrument? Could also be change from less management to more active management  
- Decline in rec effort since 2010, maybe generational shift, not at same rate for catch/landings  

Even if market driven, changes in other indicators require monitoring as they may become important drivers of landings in the future:  
- Climate trending to uncharted territory (see [Climate Risks section](#climate-and-ecosystem-productivity))  
- Stocks are shifting distribution throughout the region 

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-species-dist.R"), fig.asp=1}
```

- Some ecosystem composition and production changes (see [Stability section](#stability))  
- Engagement declining (see [Social and Cultural section](#social-and-cultural))  

## Profits 

### Indicators: revenue (a proxy for profits), with price and volume components  
Total commercial revenue continues to be driven by single species in New England; Sea scallops in GB and American lobster in the GOM.  High revenue on GB caused by high volume and price of scallops.  The cyclical nature of the indicator is caused by rotational management of the resource.  Total revenue in the GOM is above average due to high lobster prices despite low volume.  The delining trend in managed species could be tied to declines in quota or the number of stocks with rebuilding plans.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-comdat-comm-revenue.R"), fig.width = 7.5}
```

Revenue trends overall mirror those for the benthos (scallops) in GB and benthivores (lobster) in GOM over the past decade. 

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-bennet-all.R"), fig.width = 7.5, fig.asp = 0.5, echo = FALSE, message=FALSE, results='hide'}
```

### Why  
Geret to provide text 

## Recreational Opportunities 

### Indicators: Days fished, fleet diversity  
Recreational days fished have been stable over the long term, with current days fished near the long term average. Recreational fleet diversity has remained relatively stable over the time series.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-recdat-effort.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-recdat-diversity.R")}

```


### Why
Multiple drivers
-  Shift from charter party to shore based over the long term
    -  Shoreside larger component
    -  Changing species caught
(ASMFS and SAFMC managed species maintain catch diversity--next section, will break up plots)


## Stability 

### Indicators: fishery fleet and catch diversity, ecological component diversity
Commercial fishery fleet count and fleet diversity in revenue are at or near time series lows. This indicates the shift towards reliance on less species as noted in the revenue section.  Good to increase diversity or bad? Losing traditional fisheries, cascading effects on newly entered fisheries?

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-commercial-div.R"), fig.asp=1}
```

As noted [above](#recreational-opportunities), recreational fleet effort diversity is stable. However, recreational species catch diversity is increasing.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-recdat-div-catch.R")}
```

Zooplankton diversity is increasing in the GOM, while no trend is evident in GB.  Larval fish indices are steady about the mean.  Bottom trawl diversity shows a declining trend for the fall on GB with no trend in the spring or for either season in the GOM.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-zoo-diversity.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_NE.Rmd-ichthyo-diversity.R"), fig.asp=1}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_NE.Rmd-survey-shannon.R"), fig.width = 7.5}
```

 
### Why
Stability in commercial fleets reflects...  adequate capacity to respond to current fishery changes? no big shocks from management or the ecosystem so far? permitting structure that enforces similar composition? 

Declining recreational fleet effort diversity, as noted [above](#recreational-opportunities), indicates that the contraction by the party/charter boat sector continues, with shoreside angling becoming more important, by percentage.

Increase in recreational species catch diversity caused by recent increase in ASMFC and MAFMC managed species as well as decreased limits on more traditional regional species.

Increasing zooplankton diversity is driven by ... more southern species (tropicalization?) or the loss of a dominant species (Cal fin?) with everything else staying about the same?

Stable larval and adult fish diversity suggests ... the same overall number and evenness over time, but doesn't rule out species substitutions (e.g., warm-water replacing cold-water). Stable but variable (larval) can indicate changes in a dominant species. 


## Social and Cultural 

### Indicators: Commercial and recreational engagement and reliance by community
In 2020, we reported that the number of highly engaged Mid-Atlantic commercial fishing communities have declined over time, and the engagement scores have also declined in medium-highly engaged communities. Here we focus on engagement and reliance scores for individual communities, as well as social vulnerability. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-commercial-engagement.R"), fig.width = 7.5, fig.asp = 0.75}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-recreational-engagement.R"), fig.width = 7.5, fig.asp = 0.75}
```


### Why
List multiple drivers of community status: landings, permits, etc (definition of engagement and reliance), as well as drivers of social vulnerability.

Provide statement of how managers can use this status plot going forward

## Protected Species
Protected species include marine mammals (under the Marine Mammal Protection Act), endangered and threatened species (under the Endangered Species Act), and migratory birds (under the Migratory Bird Treaty Act). In the Northeast US, endangered/threatened species include Atlantic salmon, Atlantic and shortnose sturgeon, all sea turtle species, and 5 baleen whales. Fishery management objectives for protected species generally focus on reducing threats and on habitat conservation/restoration; here we report on the status of these actions as well as indicating the potential for future interactions driven by observed and predicted ecosystem changes in the Northeast US region. 

Protected species objectives include managing bycatch to remain below potential biological removal (PBR) thresholds, recovering endangered populations, and monitoring unusual mortality events (UMEs). 

### Indicators: bycatch, population (adult and juvenile) numbers, mortalities
Both harbor porpoise and gray seal bycatch indices are below current PBR thresholds, meeting management objectives. However the most recent data point for gray seals may indicate higher than allowed bycatch.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-harborporpoise.R")}
```

(Blue line here is specific to New England--did we get Mid specific estimates?)
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-grayseal.R")}
```

The North Atlantic right whale population was on a recovery trajectory until 2010, but has since declined. Reduced survival rates of adult females and diverging abundance trends between sexes have also been observed. It is estimated that there are only about 100 adult females remaining in the population. 

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-narw-abundance.R")}
```

North Atlantic right whale calf counts have also been declining. In 2018 there were zero observed new calves, and a drop in annual calves roughly mirrors the abundance decline (Fig above), however seven new calves were born in 2019. Preliminary data suggest that 2020 calf count is not 0; 12 calves have been observed as of January 2021.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-NARW-calf-abundance.R")}
```

This year saw the continuation of four Unusual Mortality Events (UMEs). Three large whale Unusual Mortality Events (UMEs) are ongoing for North Atlantic right whales, humpback whales, and minke whales. Since 2017, the total UME right whale mortalities includes 30 dead stranded whales, 9 in the US and 21 in Canada. During 2019 9 whales have been stranded in Canada and one in the US, 3 of which were determined to have occured due to vessel strike while the remainder are undetermined at this time. A UME for both gray and harbor seals was declared in 2018. 


### Why
Bycatch management measures have been implemented to maintain bycatch below thresholds. The downward trend in harbor porpoise bycatch can also be attributed to changing seasonal distributions of harbor porpoise and fisheries reducing overlap and a decrease in gill net effort. The increasing trend in gray seal bycatch is related to an increase in the gray seal population (US pup counts). 

The number of grey seals (Halichoerus grypus) in U.S. waters has risen dramatically in the last 2 decades, with few observed in the early 1990s to roughly 24,000 observed in southeastern Massachusetts in 2015. Roughly 30,000 - 40,000 gray seals were estimated in southeastern Massachusetts in 2015, using correction factors applied to seal counts visible in Google Earth imagery. As of 2016, the size of the grey seal population in Canada, which is part of the same stock as the grey seals in the U.S., was estimated to be roughly 425,000, and increasing by 4% a year. In U.S. waters, the number of pupping sites has increased from 1 in 1988 to 9 in 2019. Mean rates of increase in the number of pups born at various times since 1988 at 4 of the more data-rich pupping sites (Muskeget, Monomoy, Seal, and Green Islands) ranged from -0.2% (95%CI: -2.3 - 1.9%) to 26.3% (95%CI: 21.6 - 31.4%). These high rates of increase provide further support that seals from Canada are continually supplementing the breeding population in U.S. waters.

Strong evidence exists to suggest that the level of interaction between right whales and the combination of offshore lobster gear in the US and snow crab gear in Canada is contributing substantially to the decline of the species. Further, right whale distribution has changed since 2010. The reasons for these changes is unclear, but changes in climate and primary prey (Calanus finmarchicus) are suspected. 

UMEs are under investigation and are likely the result of multiple drivers. For all three large whale UMEs, human interaction appears to have contributed to increased mortalities, although investigations are not complete. An investigation into the cause of the seal UME so far suggests phocine distemper virus as a potential cause, although the investigation is not yet complete.

A marine mammal climate vulnerability assessment is currently underway and for Atlantic and Gulf of Mexico populations and will be reported on in future versions of this report. 

# External considerations to meeting fishery management objectives

## Climate and Ecosystem Productivity

### Climate Change Indicators: ocean currents, temperature, heatwaves, acidification
Regional ocean current indicators remain at unprecedented levels. In 2019, the Gulf Stream was at its most northern position since 1993.

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-gsi.R")}
```

In 2019, we also observed the second lowest proportion of Labrador Shelf water entering the Gulf of Maine since 1978.

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-slopewater.R")}
```

Ocean temperatures continue to warm at both the surface and bottom. Warming is not seasonally uniform, however: spring 2020 was cooler than average on portions of the shelf but generally not in New England.

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-bottom-temp.R")}
```

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-seasonal-sst-anomaly-gridded.R"), fig.width = 8, fig.asp = 1}
```

New England experienced frequent marine heatwaves of moderate to severe intensity in 2020 that extended well into December.  These included the warmest heatwave on record in GB at 4.3 degrees C above average.

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-heatwave-year-gb.R"), fig.asp=.6}
```

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-heatwave-year-gom.R"), fig.asp=.6}
```

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-heatwave.R"), fig.asp=1}
```

### Ecosystem Productivity Indicators: primary production, zooplankton, forage fish, fish condition
Increased temperatures, as reported above, can increase the rate of photosynthesis by phytoplankton (i.e. primary production).

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-pp-monthly.R"), fig.width=8, fig.asp=1}
```

Annual primary production has increased over time, primarily driven by increased productivity in the summer months. Larger than average phytoplankton blooms were observed from late fall into winter in 2020.

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-chl-weekly.R"), fig.width = 5, fig.asp = 0.9}
```

Climatology of seasonal phytoplankton size fractions confirms that the phytoplankton community in the summer is dominated by smaller size classes (pico and nano).

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-weekly-phyto-size.R"), fig.asp=.6}
```

This implies less efficient transfer of primary production to higher trophic levels

Trends in gelatinous zooplankton and krill are the same across EPUs as last year; (data was updated to 2019). There has been a long term increase in almost all regions.

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-zoo-strat-abun.R"), fig.width=8, fig.asp=.75}
```

Larger zooplankton (Cal fin) had above average abundance in 2018-2019.

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-zoo-abund.R"), fig.asp=1}
```

<!-- Trends in Cal fin overwintering stage important as forage for right whales and forage fish. -->

<!-- ```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-gb-calanus-stage.R"), fig.width=8} -->
<!-- ``` -->

<!-- ```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-gom-calanus-stage.R"), fig.width=8, fig.asp=.25} -->
<!-- ``` -->

Forage anomaly from zooplankton and ichthyoplankton data, no apparent trend in either region although the GOM anomaly is above the time series mean.

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_NE.Rmd-forage-anomaly.R"), fig.asp = .75}
```


Forage energy density updated; lower now for herring than in 1980s-90s, and lower now for  Illex squid than in the 1980s. Other species show seasonal difference but bracketing estimates from previous decades.

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-energy-density.R"), fig.width = 7.5, fig.asp = 0.5}
```


Temperature and zooplankton changes impact fish condition for different species, impacts to fisheries and markets are under investigation.
```{r, fig.width = 7.5, fig.asp = 0.5}
knitr::include_url("https://github.com/NOAA-EDAB/ecodata/raw/master/docs/images/condition_proj.png")
```


### Ecosystem Structure Indicators: distribution shifts, diversity, predators
As noted in the [Landings "Why" section above](#landings), stocks are shifting distribution throughout the region. In aggregate, fish stocks are moving northeast along the shelf and into deeper waters.  
Zooplankton diversity is increasing in the GOM, while larval fish and adult fish diversity indices are stable over time with current values near the long term average (see [Diversity Indicators section, above](#diversity)).  
New indicators for shark populations, combined with information on grey seals (see [Protected Species "Why" section, above](#protected-species)), suggests predator populations are stable to increasing in New England. Stable predator populations suggest stable predation pressure on managed species, but increasing predator populations may reflect increasing predation pressure.

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_NE.Rmd-observed-sharks.R"), fig.width = 6}
```


```{r}
#code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-hms-cpue-sharks.R"),
#sp_cat<- read.csv(here::here("data-raw/hms-mrip/hms_sp_category.csv"))
sp_cat<- read.csv("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/data-raw/hms-mrip/hms_sp_category.csv")
ecodata::hms_cpue %>% 
  filter(str_detect(Var, "SHARK")) %>% 
  rename(COMMON_POP = Var) %>% 
  left_join(sp_cat) %>% 
  group_by(Year, SP_CATEGORY) %>% 
  summarise(Value = sum(Value)) %>% 
  rename(Var = SP_CATEGORY) %>% 
  filter(!Var == "NA") %>% 
  ggplot()+
 ggplot2::annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  ggplot2::geom_point(aes(x=Year, y = Value, color = Var))+
  ggplot2::geom_line(aes(x=Year, y = Value, color = Var))+
  #ggplot2::facet_wrap(~Var, scales = "free")+
  ggplot2::ggtitle("HMS POP SHARK CPUE")+
  ggplot2::ylab("Number per Haul")+
  ecodata::theme_ts()
```

As noted above in the Protected Species section, gray seal populations are increasing. Current information suggests that gray seals eat primarily sand lance, hakes and flatfish, and squids, while harbor seals consume a variety of groundfish (hakes, cod, haddock, flatfish), redfish, herring and squids, however much of this information comes from juvenile animals and more research is needed on animals at other life stages. Additional analysis of gray and harbor seal diet is currently underway at the NEFSC using a variety of techniques (analysis of stomach contents, fatty acids, and DNA). This information can eventually be coupled with estimates of population abundance and consumption rates to estimate total biomass removals of fish due to pinniped predation.

### Habitat Climate Vulnerability
New habitat climate vulnerability analysis links several NEFMC species to several highly vulnerable nearshore habitats.  In particular, winter flounder, a species deemed highly vulnerable to climate change itself, was found to be highly dependent on vulnerable habitats throughout New England and the Mid-Atlantic. Details on highly vulnerable habitats with linkages to a variety of species, including which life stages have different levels of dependence on a particular habitat, are available in a detailed table^[https://noaa-edab.github.io/ecodata/Hab_table].

### Implications

Climate-managed species links

2012 still warmest on record, now papers published on impacts to give insight into what the implications of heatwaves can be
Lobster was a big one--timing of fishing, markets, etc
Cobia habitat shifts (and other species)--seen off RI now

Shifting species distributions alter both species interactions and fishery interactions. In particular, shifting species distributions can alter expected management outcomes from spatial allocations based on historical distributions. 
Climate affects ecosystem productivity, with warming waters increasing the rate of photosynthesis at the base of the food web. However, increased summer production in New England may not translate to increased fish biomass because smaller phytoplankton dominate in this season, and trends in ecosystem components are not consistently in a single direction. While krill and large gelatinous zooplankton are increasing over time, smaller zooplankton are periodically shifting abundance between the larger, more nutritious Cal fin and smaller bodied copepods with no apparent overall trend.  Forage species are difficult to survey, but a new index based on ichthyoplankton suggests high interannual variability in abundance. Forage species nutrition changes seasonally in response to ecosystem conditions, with apparent declines in energy density for Atlantic herring and Illex squid relative to the 1980s, but similar energy density for other forage species. Some of these factors are now being linked to fish condition.

Fish Condition Links to Environmental Drivers

Generalized Additive Models (GAMs) were used to test how measures of fishing pressure, stock abundance and individual environmental variables performed in explaining the changes of fish condition (fatness) over time. Some species such as Acadian redfish, butterfish and winter flounder were more affected by fishing pressure and stock size, whereas other species such as weakfish, windowpane and American plaice may be more affected by local bottom temperatures and zooplankton. These relationships can potentially provide insights on which species may be more vulnerable to environmental changes such as climate change, as well as what biomass changes may be expected from certain species given current environmental conditions. Correlations were examined between environmental drivers, and as expected there were strong temperature correlations between seasons as well as correlations between temperature and zooplankton indices. Planned future work includes building full GAM models for each fish species, and linking fish condition to socio-economic models to assess whether fish condition impacts the value generated by that species.
 
Socio-Economic Models
 
Economic theory and empirical analyses have highlighted that many factors can affect the price of fish, including the total quantity of fish in the market (sometimes including internationally), increased demand around holidays, time the fish was in storage, and other issues that either affect the quality of the fish or the amount of fish available for purchase. We plan on empirically exploring whether fish condition is a quality metric people care about. Understanding whether people care about fish condition will help us more holistically understand the impacts of condition change on society, if any.

## Other Ocean Uses: Offshore Wind

### Indicators: revenue in lease areas, development timeline, survey overlap
Commercial fishery revenue in proposed offshore wind lease areas, including the New York Bight Call Areas, ranges from 2-3% of total revenue for NEFMC managed species.

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_NE.Rmd-wind-revenue.R"), fig.width = 6}
```

Offshore wind projects may cover more than 1.5 million acres by 2029. Although this rapid build out will be more impactful to the Mid-Atlantic than New England which only has some lease areas in RI and MA.  It is possible floating offshore technologies may be utilized in GOM in the future.


```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-wind-dev-speed0.R")}
```

Scientific survey area overlap with offshore wind lease areas ranges from 2-25% across ecosystem, shellfish, fish, and protected species surveys.  Once again, the majority of this overlap will occur in the Mid-Atlantic but may have implications on assessments for NEFMC managed species.

```{r}
knitr::include_url("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/docs/images/SurveysWindFarmAreas_20201229.jpg")
```

### Implications
Fishery revenue could be displaced to other areas or foregone if all wind projects are built.  
Displaced fishing effort can alter fishing methods and change habitat, species (managed and protected), and fleet interactions.  
Right whales may be displaced and could face possible prey impacts via changed oceanography.  
Current plans for rapid buildout of offshore wind in a patchwork of areas spreads the impacts differentially throughout the region. 

```{r, fig.asp=1.5}
knitr::include_url("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/docs/images/MidAtlantic_2021119-01.jpg")
```

Offshore wind will pose a risk to scientific data collection for surveys and increased uncertainty for management decision making across physical, low trophic, fish, and protected species.




# Contributors

**Editors** (NOAA NMFS Northeast Fisheries Science Center, NEFSC): Sean Lucey, Sarah Gaichas, Kimberly Bastille, Geret DePiper, Kimberly Hyde, Scott Large, Chris Orphanides 

**Contributors** (NEFSC unless otherwise noted): Andy Beet, Ruth Boettcher (Virginia Department of Game and Inland Fisheries), Zhuomin Chen (Woods Hole Oceanographic Institute), Doug Christel (GARFO), Patricia Clay, Lisa Colburn, Jennifer Cudney (NMFS Atlantic HMS Management Division), Tobey Curtis (NMFS Atlantic HMS Management Division), Geret DePiper, Michael Fogarty, Paula Fratantoni, Kevin Friedland, Sarah Gaichas, Avijit Ben Galuardi (GARFO), Gangopadhyay (School for Marine Science and Technology, University of Massachusetts Dartmouth), James Gartland (Virginia Institute of Marine Science), Glen Gawarkiewicz (Woods Hole Oceanographic Institution), Sean Hardison, Kimberly Hyde, John Kosik, Steve Kress (National Audubon Society’s Seabird Restoration Program), Young-Oh Kwon (Woods Hole Oceanographic Institute), Andrew Lipsky, Sean Lucey, Don Lyons (National Audubon Society’s Seabird Restoration Program), Chris Melrose, Ryan Morse, Kimberly Murray, Chris Orphanides, Richard Pace, Charles Perretti, Grace Saba (Rutgers), Vincent Saba, Angela Silva, Laurel Smith, Talya tenBrink (GARFO), John Walden, Harvey Walsh, Mark Wuenschel.  

\newpage 

# Document Orientation

The figure format is illustrated in Fig \ref{fig:docformat}a. Trend lines are shown when slope is significantly different from 0 at the p < 0.05 level. An orange line signifies an overall positive trend, and purple signifies a negative trend. To minimize bias introduced by small sample size, no trend is fit for < 30 year time series. Dashed lines represent mean values of time series unless the indicator is an anomaly, in which case the dashed line is equal to 0. Shaded regions indicate the past ten years. If there are no new data for 2018, the shaded region will still cover this time period. The spatial scale of indicators is either coastwide, New England states (Connecticut, Rhode Island, Massachusetts, New Hampshire, and Maine), or at one of the two Ecosystem Production Unit (EPU, Fig. \ref{fig:docformat}b) levels in the region, Georges Bank (GB) or Gulf of Maine (GOM).

```{r docformat, fig.cap = "Document orientation. a. Key to figures. b.The Northeast Large Marine Ecosystem.",  fig.width = 8, fig.height = 2.5}
#fig.subcap= c('Key to figures.', 'The Northeast Large Marine Ecosystem.'), out.width = '.49\\linewidth', fig.show="hold" 
# FIgure orientation subfigure
m <- 0.1
x <- 1985:2018
y <-  m*x + rnorm(30, sd = 0.35)

data <- data.frame(x = x,
                  y = y)

#Define constants for figure plot
x.shade.max <- max(x)
x.shade.min <- x.shade.max - 9 
hline = mean(y)

#Plot series with trend 
psample <- ggplot2::ggplot(data = data,aes(x = x, y = y)) +
  #Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  geom_point(size = pcex) +
  scale_color_manual(aesthetics = "color")+
  guides(color = FALSE) +
  geom_hline(aes(yintercept = hline),
              size = hline.size,
             alpha = hline.alpha,
             linetype = hline.lty)+
  geom_line() +
  geom_gls() +
  scale_y_continuous(labels = function(l){trans = l / 1000})+
  scale_x_continuous(breaks = seq(1985, 2015, by = 5), expand = c(0.01, 0.01)) +
  ylab(expression("Invented Index, 10"^3*"widgets")) +
  xlab("Time") +
  labs(tag = "a")  +
  theme_ts()


# EPU map subfigure
# Set lat/lon window for map
xmin = -78
xmax = -65
ymin = 36
ymax = 45
xlims <- c(xmin, xmax)
ylims <- c(ymin, ymax)

#CRS
crs <- "+proj=longlat +lat_1=35 +lat_2=45 +lat_0=40 +lon_0=-77 +x_0=0 +y_0=0 +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"

#Specify data frame with lat/lon locations for labels
epu_labels <- data.frame(EPU = c("Mid-Atlantic\n Bight",
                                 "Gulf of Maine",
                                 "Georges Bank"),
                         latitude = c(40,42.85,41),
                         longitude = c(-72.7,-69,-68.5))

#Map of NE LME
epumap <- ggplot() +
  geom_sf(data = coast, size = map.lwd) +
  geom_sf(data = epu_sf, fill = "transparent", size = map.lwd) +
  coord_sf(crs = crs, xlim = xlims, ylim = ylims) +
  geom_text(data = epu_labels, aes(x = longitude,
                                    y = latitude,
                                    label = EPU),
            size = 1.7)+
  theme_map() +
  scale_x_continuous(breaks = seq(-78, -65, by = 4), expand = c(0.01, 0.01)) +
  xlab("Longitude") +
  ylab("Latitude") +
  labs(tag = "b")

psample + plot_spacer() + epumap + plot_layout(ncol = 3, widths=c(3,.5,4.5))

```

Fish and invertebrates are aggregated into similar feeding categories (Table \ref{tab:species-groupings}) to evaluate ecosystem level trends in predators and prey. 

```{r species-groupings, warning=F}

# new table with all species listed by management entity
df <- ecodata::species_groupings %>%
  dplyr::select(SOE.20, COMNAME, Fed.Managed) %>%
  filter(SOE.20 != "Other") %>%
  distinct() %>%
  group_by(SOE.20, Fed.Managed) %>%
  summarize_all(funs(paste(na.omit(.), collapse = ", "))) %>%
  spread(Fed.Managed, COMNAME) %>%
  arrange(factor(SOE.20, levels = c("Apex Predator", "Piscivore", "Planktivore", "Benthivore", "Benthos")))
df<-df[c(1,3,2,4,5)] %>%
  mutate_all(tolower)


knitr::kable(df, booktabs = TRUE, caption = 'Feeding guilds and management bodies.', 
             col.names = c("Guild", "MAFMC", "Joint", "NEFMC", "State or Other")) %>%
  kable_styling(font_size=10, latex_options=c("repeat_header", "scale_down", "hold_position")) %>%
  row_spec(0,bold=TRUE) %>%
  column_spec(1, width="2cm") %>%
  column_spec(2, width="4cm") %>%
  column_spec(3, width="2cm") %>%
  column_spec(4, width="5cm") %>%
  column_spec(5, width="6cm") %>%
  #column_spec(3, width="7.5cm") #%>%
  collapse_rows(columns = 1, latex_hline = "major", valign = "middle")

```


\newpage 

# References
